Make any stupid plot you want

### 2021-05-14 draft/sections/intro.tex
color coding to clean up explainations

Most information visualization software design patterns are tuned to specific data structures and domains, as categorized by Heer and Agrawala  \cite{HeerSoftware2006}. For users who generally work in one domain--such relational tables (\autoref{fig:intro:table}), images (\autoref{fig:intro:image}), or graphs (\autoref{fig:intro:graph})--well defined data space (and corresponding visual space  \cite{chiTaxonomyVisualizationTechniques2000}) often yields a coherent user experience \cite{heerDeclarative2010}. This coherent experience is often not extensible; developers who want to build new visualizations on top of these libraries must work around the existing assumptions, sometimes in ways that break the architecture model of the libraries. For example tools influenced by APT that assume that data is a relational table integrate computation into the visualization pipeline. This is a wide array of tools, including Tableau \cite{StoltePolaris2002,hanrahanVizQL2006,MackinlayShowme2007} and the Grammar of Graphics \cite{wilkinsonGrammarGraphics2005} inspired ggplot \cite{wickhamGgplot2ElegantGraphics2016a}, protovis \cite{bostockProtoviz2009}, vega \cite{satyanarayanDeclarativeInteractionDesign2014} and altair \cite{vanderplasAltairInteractiveStatistical2018}. Since these libraries represent data as a table, and computations on tables are well defined \cite{ullmanFirstCourseDatabase2008}, these libraries implement computation as part of the visualization process. This is also true of tools that support images, such as the biology oriented ImageJ \cite{schneiderNIHImageImageJ2012} and  Napari \cite{nicholas_sofroniew_2021_4533308} or the digital humanities ImageJ macro ImagePlot \cite{studiesCulturevisImageplot2021}. While these tools often have some support for visualizing non image components of the data, the architecture is oriented towards building plugins into the existing system \cite{WritingPlugins} where the image is the core data structure. Tools like Gephi \cite{bastianGephiOpenSource2009}, Graphviz \cite{ellsonGraphvizOpenSource2002}, and Networkx \cite{HagbergExploringNetwork2008} are used to  visualize and manipulate graphs. As with tables and images, extending network libraries to work with other types of data either requires breaking the libraries internal model of how data is structured and what data transformations are allowed or implementing a model for other types of data structures alongside the network model. Our model aims to identify which computations are manipulations of the data and which are necessary for the visual encoding; for example data is often aggregated for a bar chart but it is not required, while a boxplot by definition requires computing distribution statistics \cite{wickham40YearsBoxplots2011}. Disentangling the computation from the visual transforms allows us to determine whether the visualization library should implement computations or if they should be implemented in data space.

>>>>

Most information visualization software design patterns are tuned to specific data structures and domains, as categorized by Heer and Agrawala  \cite{HeerSoftware2006}. For users who generally work in one domain--such relational tables (\autoref{fig:intro:table}), images (\autoref{fig:intro:image}), or graphs (\autoref{fig:intro:graph})--well defined data space (and corresponding visual space  \cite{chiTaxonomyVisualizationTechniques2000}) often yields a coherent user experience \cite{heerDeclarative2010}. This coherent experience is often not extensible; developers who want to build new visualizations on top of these libraries must work around the existing assumptions, sometimes in ways that break the architecture model of the libraries. For example tools influenced by APT that assume that data is a relational table integrate computation into the visualization pipeline. This is a wide array of tools, including Tableau \cite{StoltePolaris2002,hanrahanVizQL2006,MackinlayShowme2007} and the Grammar of Graphics \cite{wilkinsonGrammarGraphics2005} inspired ggplot \cite{wickhamGgplot2ElegantGraphics2016a}, protovis \cite{bostockProtoviz2009}, vega \cite{satyanarayanDeclarativeInteractionDesign2014} and altair \cite{vanderplasAltairInteractiveStatistical2018}. Since these libraries represent data as a table, and computations on tables are well defined \cite{ullmanFirstCourseDatabase2008}, these libraries implement computation as part of the visualization process. This is also true of tools that support images, such as the biology oriented ImageJ \cite{schneiderNIHImageImageJ2012} and  Napari \cite{nicholas_sofroniew_2021_4533308} or the digital humanities ImageJ macro ImagePlot \cite{studiesCulturevisImageplot2021}. While these tools often have some support for visualizing non image components of the data, the architecture is oriented towards building plugins into the existing system \cite{WritingPlugins} where the image is the core data structure. Tools like Gephi \cite{bastianGephiOpenSource2009}, Graphviz \cite{ellsonGraphvizOpenSource2002}, cytoscape \cite{shannonCytoscapeSoftwareEnvironment2003}, and Networkx \cite{HagbergExploringNetwork2008} are used to  visualize and manipulate graphs. As with tables and images, extending network libraries to work with other types of data either requires breaking the libraries internal model of how data is structured and what data transformations are allowed or implementing a model for other types of data structures alongside the network model. Our model aims to identify which computations are manipulations of the data and which are necessary for the visual encoding; for example data is often aggregated for a bar chart but it is not required, while a boxplot by definition requires computing distribution statistics \cite{wickham40YearsBoxplots2011}. Disentangling the computation from the visual transforms allows us to determine whether the visualization library should implement computations or if they should be implemented in data space.
